---
title: "Workshop 1-Into the programing"
author: "Agustinus Satrio Supoyo"
date: "2026-05-20"
output:
  html_document: default
  pdf_document: default
---
## Introduction

This workshop introduced the fundamental skills needed to begin programming in R. I learned how to navigate RStudio, write and run scripts, create objects, use functions, understand data types, and organise files within an R Project.

These foundations are important for marine science because modern datasets—such as coral surveys, satellite observations, acoustic records, and animal tracking data—require clear, reproducible, and well-organised analytical workflows.

### Working Directory

```{r working-directory}
getwd()
```

### Topics Covered

- Basic calculations
- Comments and code style
- Objects and assignment
- Object naming rules
- Functions and arguments
- Creating text with `paste()`
- Debugging
- Data types
- Vectors and lists
- Data frames
- Missing data
- Tibbles
- Packages
- R Projects and file organisation

## 1. Basic Calculations

R can be used as a calculator.

```{r basic-calculations}
24.5 + 1.2
32.0 / 4
2 + 1
6 * 2
```

The `*` symbol represents multiplication.

The colon operator `:` creates a sequence of numbers.

```{r number-sequence}
1:30
```

Numbers such as `[1]`, `[9]`, or `[17]` in the Console indicate the index of the first value displayed on each line.

## 2. Comments and Code Style

Comments begin with the `#` symbol. R ignores text written after `#`.

```{r comments-and-style}
# Calculate the total area surveyed

quadrat_area_m2 <- 0.25
number_of_quadrats <- 16

total_area_surveyed <-
  quadrat_area_m2 * number_of_quadrats

total_area_surveyed
```

Comments can be used to explain code, organise scripts, and record useful notes.

## 3. Objects and Assignment

The assignment operator `<-` stores information inside an object.

```{r objects-and-assignment}
coral_count <- 42

fish_lengths <- c(
  124,
  152,
  98,
  221,
  146
)

coral_count
fish_lengths
```

Objects can be used in calculations.

```{r manipulating-objects}
coral_count + 1
coral_count + coral_count
```

R is case-sensitive.

```{r case-sensitivity}
coral_count <- 42
Coral_Count <- 1

coral_count
Coral_Count

coral_count + Coral_Count
```

## 4. Object Naming Rules

Object names should be clear, descriptive, and consistent.

```{r valid-object-names}
age_in_years <- 25
transect_depth_m <- 12.5
bleaching_present <- TRUE
```

Object names cannot normally begin with a number, contain spaces, or begin with unsupported special symbols.

The following examples are displayed but not run.

```{r invalid-object-names, eval=FALSE}
01_age <- 25
!_age <- 25
coral count <- 25
```

Spaces can technically be used with backticks.

```{r backtick-object-name}
`coral count` <- 25

`coral count`
```

However, underscores are usually clearer.

```{r recommended-object-name}
coral_count <- 25
```

## 5. Functions and Arguments

Functions perform specific operations in R.

### Rounding Numbers

```{r rounding-functions}
years_old <- 25.7

round(years_old)
floor(years_old)
```

The second argument of `round()` controls the number of decimal places.

```{r rounding-decimal-places}
years_old <- 25.765

round(years_old, 2)
round(years_old, 1)
```

Function help and arguments can be viewed using:

```{r function-help, eval=FALSE}
?round
args(round)
```

## 6. Using `paste()` to Create Sentences

The `paste()` function combines text and object values.

```{r paste-sentence}
years_old <- 25.765
rounded_age <- round(years_old, 2)

sentence <- paste(
  "Average colony age is",
  rounded_age,
  "years old."
)

print(sentence)
```

The value can also be rounded to one decimal place.

```{r paste-one-decimal}
rounded_age <- round(years_old, 1)

paste(
  "Average colony age is",
  rounded_age,
  "years old."
)
```

## 7. Debugging

Debugging is the process of identifying and correcting errors in code.

The following example contains a spelling error and is not run during knitting.

```{r debugging-error, eval=FALSE}
quadrat_area_m2 <- 0.25
number_of_quadrats <- 16

total_area_surveyed <-
  quadrat_area_m2 * number_of_quadrats

print(total_area_surveyd)
```

The corrected version is:

```{r debugging-corrected}
quadrat_area_m2 <- 0.25
number_of_quadrats <- 16

total_area_surveyed <-
  quadrat_area_m2 * number_of_quadrats

print(total_area_surveyed)
```

Incomplete or incorrect syntax can also produce errors.

```{r syntax-errors, eval=FALSE}
6 *

6 % 2
```

The correct multiplication command is:

```{r correct-syntax}
6 * 2
```

## 8. Data Types

R uses different data types to determine which operations can be performed.

### Character

```{r character-data}
site_name <- "Heron_Island"

class(site_name)
str(site_name)
```

### Numeric

```{r numeric-data}
transect_depth_m <- 12.5

class(transect_depth_m)
str(transect_depth_m)
```

### Integer

```{r integer-data}
turtle_count <- 5L

class(turtle_count)
str(turtle_count)
```

### Logical

```{r logical-data}
bleaching_present <- TRUE

class(bleaching_present)
str(bleaching_present)
```

## 9. Vectors and Lists

A vector stores multiple values of the same data type.

### Numeric Vector

```{r numeric-vector}
fish_lengths <- c(
  124,
  152,
  98,
  221,
  146
)

fish_lengths
class(fish_lengths)
length(fish_lengths)
```

### Character Vector

```{r character-vector}
coral_species <- c(
  "Porites",
  "Acropora",
  "Montastrea"
)

coral_species
class(coral_species)
```

### Mixed Values in a Vector

Atomic vectors can only contain one data type. R converts all values to a common type.

```{r vector-coercion}
notes_vector <- c(
  "Acropora",
  27.5,
  TRUE
)

notes_vector
class(notes_vector)
```

### Lists

A list can store several different data types.

```{r mixed-list}
notes_list <- list(
  "Acropora",
  27.5,
  TRUE
)

notes_list
class(notes_list)
```

Individual list elements can be accessed using double square brackets.

```{r access-list-elements}
notes_list[[1]]
notes_list[[2]]
notes_list[[3]]
```

## 10. Data Frames

A data frame stores rectangular data in rows and columns.

```{r create-data-frame}
my_dataframe <- data.frame(
  no = c(1, 2, 3),
  species = c(
    "Plectropomus",
    "Scarus",
    "Pomacentrus"
  ),
  recorded = c(
    TRUE,
    FALSE,
    TRUE
  )
)

my_dataframe
```

The `str()` function displays the structure of the data frame.

```{r inspect-data-frame}
str(my_dataframe)
```

A column can be converted to a factor when it represents categorical data.

```{r convert-column-to-factor}
my_dataframe$no <- as.factor(
  my_dataframe$no
)

str(my_dataframe)
```

## 11. Missing Data

The value `NA` represents missing or unavailable data.

```{r create-missing-data}
coral_height_cm <- c(
  170,
  NA,
  180
)

coral_height_cm
```

The `is.na()` function identifies missing values.

```{r check-missing-data}
is.na(coral_height_cm)
```

The `na.omit()` function removes missing values.

```{r remove-missing-data}
na.omit(coral_height_cm)
```

## 12. Tibbles

A tibble is a modern version of a data frame used extensively in the tidyverse.

### Load the `tibble` Package

```{r load-tibble}
library(tibble)
```

### Create a Tibble

```{r create-tibble}
my_tibble <- tibble(
  no = c(1, 2, 3),
  species = c(
    "Plectropomus",
    "Scarus",
    "Pomacentrus"
  ),
  recorded = c(
    TRUE,
    FALSE,
    TRUE
  )
)

my_tibble
```

## 13. Installing and Loading Packages

Packages contain functions created by other programmers and researchers.

Packages should normally be installed once through the Console.

```{r install-package, eval=FALSE}
install.packages("tidyverse")
```

Packages must be loaded during each new R session.

```{r load-package, eval=FALSE}
library(tidyverse)
```

Individual packages can also be loaded separately.

```{r load-individual-packages, eval=FALSE}
library(dplyr)
library(ggplot2)
```

Multiple packages can be installed together.

```{r install-multiple-packages, eval=FALSE}
install.packages(
  c(
    "ggplot2",
    "tidyr",
    "dplyr",
    "terra"
  )
)
```

## 14. R Projects and File Organisation

The `getwd()` function displays the current working directory.

```{project-working-directory}
getwd()
```

R Projects provide a reliable way to organise files and use relative file paths.

Hardcoded working directories should be avoided.

```{r avoid-setwd, eval=FALSE}
setwd(
  "C:/Users/Name/Desktop/MyDocuments/Class2026/Data"
)
```

A recommended project structure is:

```text
MB5370_Module_1_Coding/
│
├── MB5370_Module_1_Coding.Rproj
├── data/
├── code/
├── output/
└── docs/
```

The folders can be created using:

```{r create-project-folders, eval=FALSE}
dir.create("data")
dir.create("code")
dir.create("output")
dir.create("docs")
```

The folders are used for:

- `data/`: raw and unedited data
- `code/`: R scripts and data-processing code
- `output/`: figures, tables, and processed files
- `docs/`: rendered reports and final documents