---
title: "Workshop 02"
author: "Agustinus Satrio Supoyo"
date: "2026-07-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Load Packages 
```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
```
##ggplot

#| Geom               | Purpose      |
#| `geom_point()`     | Scatter plot |
#| `geom_line()`      | Line chart   |
#| `geom_bar()`       | Bar chart    |
#| `geom_boxplot()`   | Boxplot      |
#| `geom_histogram()` | Histogram    |
#| `geom_smooth()`    | Trend line   |


```{r}
mpg
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

#Aesthetic mappings
aes() #used to connect variable to visual element
Visual elements include:
- `colour`
- `size`
- `shape`
- `alpha`
- `linetype`
- `group`


### Change Colour Based on Class

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

### Set a Fixed Colour
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy),colour="blue")
```


### Change Point Size by Class
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

### Change Point Shape by Class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

### Change Transparency
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

### Colour Based on a Logical Condition
```{r}
ggplot(data = mpg) +
  geom_point( mapping = aes( x = displ,y = hwy,colour = displ < 5))
```

##Troubleshooting
```{r}
ggplot(data = mpg)+ 
 geom_point(mapping = aes(x = displ, y = hwy)) 
# the + should be on top line
```

```{r}
ggplot(data = mpg)+
 geom_point(mapping = aes(x = displ, y = hwy)) 
```

##Faceting (Panel Plot)
Faceting divides a plot into several smaller panels.

### facet_wrap()
```{r}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2)
```

### facet_grid()
```{r}
#Face grid
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl)
```

## Smooth Line
### Add a Smooth Line
```{r}
ggplot(data=mpg)+
  geom_smooth(mapping=aes(x=displ,y=hwy))
```

### Combine Points and Smooth Line
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line
```

### Change Line Type by Drive Type
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

### Group Data by Drive Type
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

### Change Line Colour by Group
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, colour = drv),
    show.legend = FALSE
  )
```

##Combining Multiple Geoms
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

##Using Global Mappings
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth() #to avoid repeating the same aes()

```

## Layer-Specific Mappings
```{r}
 ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, colour = drv),
    show.legend = FALSE,
  )

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(colour = class)) + 
  geom_smooth()

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(colour = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)


#Code 1 — Global Mapping
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
#Code 2 — Local Mapping
ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))



```

##Plotting Statistics
###create the bar chart
```{r}

ggplot(data = diamonds) +
  geom_bar(aes(x = cut))
```

### Use stat_count()
```{r}
ggplot(data = diamonds) +
  stat_count(aes(x = cut)) #the same way
```

## Overriding ggplot2 Defaults
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```

##override a default mapping from transformed variables to aesthetics
```{r}

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
#> Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.


```

##Plotting statistical details
```{r}
#using stat_summary().
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )

```

##aethetic adjustment fill and color
### Colour the Bar Outlines
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
```

### Fill Bars by Cut
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

### Fill Bars by Clarity

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

## Position Adjustments
###To alter transparency (alpha)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
```


###To color the bar outlines with no fill color
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```


###Position = "fill" works like stacking, but scales each bar, so that all bars have the same total height (100%).
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```



###position = "dodge" places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```


###Position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```


