---
title: "Workshop 4"
author: "Tio"
date: "2026-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

#
## Exercise: AI-Driven Data Visualisation Challenge

In this exercise, you will revisit the data visualisation skills introduced in Workshop 2. Instead of writing all code manually, you will use inline comments to guide GitHub Copilot.

The exercise uses the `palmerpenguins` dataset, which contains biological and environmental measurements for three penguin species from the Palmer Archipelago in Antarctica.


# ------------------------------------------------------------
# 1. Load Required Packages
# ------------------------------------------------------------


```{r}
library(tidyverse)
library(palmerpenguins)
library(purrr)
```

# ------------------------------------------------------------
# 2. Data Visualisation with ggplot2
# ------------------------------------------------------------

# Scatter plot of penguin bill length and bill depth


```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    colour = species
  )
) +
  geom_point(
    size = 3,
    na.rm = TRUE
  ) +
  theme_minimal() +
  labs(
    title = "Bill Length vs Bill Depth of Penguins",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    colour = "Species"
  )
```

# Scatter plot of body mass and flipper length
# A separate linear regression line is fitted for each species
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = flipper_length_mm,
    y = body_mass_g,
    colour = species
  )
) +
  geom_point(
    size = 3,
    na.rm = TRUE
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    na.rm = TRUE
  ) +
  theme_minimal() +
  labs(
    title = "Body Mass vs Flipper Length of Penguins",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    colour = "Species"
  )

```

# ------------------------------------------------------------
# 3. Optional AI Packages
# ------------------------------------------------------------

# Run these installation commands only once if needed
# install.packages("ellmer")
# install.packages("chattr")

# Launch the chattr application
# chattr::chattr_app()


# ------------------------------------------------------------
# 4. Conditional Statements
# ------------------------------------------------------------

# Example: checking sea surface temperature
```{r}
sst <- 30.2

if (sst > 29.5) {
  print("Warning: Marine heatwave threshold exceeded!")
} else {
  print("SST remains within baseline parameters.")
}
```

# ------------------------------------------------------------
# 5. Vectorised Conditional Functions
# ------------------------------------------------------------

# Sample coral monitoring data
```{r}
coral_monitoring <- tibble(
  site = c("Site_A", "Site_B", "Site_C"),
  depth_m = c(12, 35, 8)
)
```


# Classify reef depth using if_else()
```{r}
coral_monitoring <- coral_monitoring %>%
  mutate(
    zone = if_else(
      depth_m > 30,
      "Deep Reef",
      "Shallow Reef"
    )
  )

```

# Classify reef depth using case_when()
```{r}
coral_monitoring <- coral_monitoring %>%
  mutate(
    reef_category = case_when(
      depth_m < 10 ~ "Lagoon / Flats",
      depth_m <= 30 ~ "Crest / Slope",
      depth_m > 30 ~ "Mesophotic / Deep",
      TRUE ~ "Unclassified"
    )
  )

coral_monitoring
```


# ------------------------------------------------------------
# 6. Salinity Classification
# ------------------------------------------------------------

```{r}
marine_stations <- tibble(
  salinity = c(35, 28, 32, 12)
)

marine_stations <- marine_stations %>%
  mutate(
    environment_type = case_when(
      salinity < 15 ~ "Estuarine",
      salinity >= 15 & salinity <= 30 ~ "Brackish",
      salinity > 30 ~ "Marine",
      TRUE ~ "Unclassified"
    )
  )

marine_stations


```

# ------------------------------------------------------------
# 7. Basic for-loops
# ------------------------------------------------------------

# Process climate data for multiple years
```{r}
for (year in 2020:2024) {
  print(
    paste(
      "Processing climate data for year:",
      year
    )
  )
}
```

# Print transect length information
```{r}
transect_lengths <- c(50, 100, 25, 75)

for (i in seq_along(transect_lengths)) {
  print(
    paste(
      "Transect number",
      i,
      "measures",
      transect_lengths[i],
      "metres."
    )
  )
}

```

# ------------------------------------------------------------
# 8. Saving Loop Results
# ------------------------------------------------------------

# Calculate square roots using a for-loop
```{r}
site_areas <- c(144, 400, 625)

square_root_results <- numeric(length(site_areas))

for (i in seq_along(site_areas)) {
  square_root_results[i] <- sqrt(site_areas[i])
}

square_root_results
```


# Calculate square roots using purrr::map_dbl()
```{r}
mapped_results <- map_dbl(
  site_areas,
  sqrt
)

mapped_results
```

# ------------------------------------------------------------
# 9. Iteration over Data Groups
# ------------------------------------------------------------

# Split the iris dataset by species and summarise each group
```{r}
iris_summaries <- iris %>%
  split(.$Species) %>%
  map(summary)

iris_summaries
```

# ------------------------------------------------------------
# 10. Comparing a for-loop with purrr
# ------------------------------------------------------------

```{r}
fish_counts <- list(
  c(10, 12, 15, 11),
  c(8, 9, 7, 10, 12),
  c(5, 6, 4)
)
```

# Calculate mean fish counts using a for-loop
```{r}
mean_counts_for <- numeric(length(fish_counts))

for (i in seq_along(fish_counts)) {
  mean_counts_for[i] <- mean(fish_counts[[i]])
}

mean_counts_for

```

# Calculate mean fish counts using purrr::map_dbl()
```{r}
mean_counts_purrr <- map_dbl(
  fish_counts,
  mean
)

mean_counts_purrr
```

# ------------------------------------------------------------
# 11. Custom Function: Coral Mortality
# ------------------------------------------------------------
```{r}
calculate_coral_mortality <- function(
    initial_count,
    surviving_count
) {
  
  # Check that the initial count is greater than zero
  if (initial_count <= 0) {
    stop("Initial count must be greater than zero.")
  }
  
  # Check that surviving count is not negative
  if (surviving_count < 0) {
    stop("Surviving count cannot be negative.")
  }
  
  # Check that surviving count does not exceed initial count
  if (surviving_count > initial_count) {
    stop("Surviving count cannot exceed initial count.")
  }
  
  mortality_rate <- (
    initial_count - surviving_count
  ) / initial_count
  
  return(mortality_rate)
}
```


# Test the coral mortality function
```{r}
coral_mortality_result <- calculate_coral_mortality(
  initial_count = 120,
  surviving_count = 84
)

coral_mortality_result
```

# ------------------------------------------------------------
# 12. Student Exercise 4
# Convert Celsius to Fahrenheit
# ------------------------------------------------------------

```{r}
convert_temp_c_to_f <- function(celsius) {
  
  # Check that the input is a single numeric value
  if (
    !is.numeric(celsius) ||
    length(celsius) != 1 ||
    is.na(celsius)
  ) {
    stop("Celsius must be a single non-missing numeric value.")
  }
  
  # Stop if the temperature is below absolute zero
  if (celsius < -273.15) {
    stop(
      "Temperature cannot be below absolute zero (-273.15°C)."
    )
  }
  
  fahrenheit <- (celsius * 9 / 5) + 32
  
  return(fahrenheit)
}


```


# Test with a valid temperature
```{r}
valid_temperature <- convert_temp_c_to_f(
  celsius = 25
)

valid_temperature

```

# Test with an invalid temperature
# tryCatch() allows the script to continue after the error
```{r}
invalid_temperature_test <- tryCatch(
  {
    convert_temp_c_to_f(
      celsius = -300
    )
  },
  error = function(error_message) {
    print(error_message$message)
  }
)

```


# ------------------------------------------------------------
# 13. Advanced Challenge 1
# Nested Multi-Vector Iteration
# ------------------------------------------------------------

```{r}
counts <- c(15, 24, 8, 42)

scaling_factors <- c(1.2, 0.8, 2.5, 1.1)

```


# Multiply each count by its matching scaling factor
```{r}
scaled_abundance <- map2_dbl(
  counts,
  scaling_factors,
  ~ .x * .y
)

scaled_abundance
```

# ------------------------------------------------------------
# 14. Advanced Challenge 2
# Robust Data-Wrangling Pipeline Function
# ------------------------------------------------------------

```{r}
clean_and_classify_survey <- function(
    data,
    column_name,
    threshold
) {
  
  # Check that data is a data frame or tibble
  if (!is.data.frame(data)) {
    stop(
      "The 'data' argument must be a data frame or tibble."
    )
  }
  
  # Check that the column name is a single character string
  if (
    !is.character(column_name) ||
    length(column_name) != 1 ||
    is.na(column_name)
  ) {
    stop(
      "The column name must be provided as a single character string."
    )
  }
  
  # Check that the specified column exists
  if (!column_name %in% names(data)) {
    stop(
      paste0(
        "Column '",
        column_name,
        "' was not found in the supplied data frame."
      )
    )
  }
  
  # Check that the selected column is numeric
  if (!is.numeric(data[[column_name]])) {
    stop(
      paste0(
        "Column '",
        column_name,
        "' must contain numeric values."
      )
    )
  }
  
  # Check that the threshold is a valid numeric value
  if (
    !is.numeric(threshold) ||
    length(threshold) != 1 ||
    is.na(threshold)
  ) {
    stop(
      "The threshold must be a single non-missing numeric value."
    )
  }
  
  # Remove missing values and classify each observation
  cleaned_data <- data %>%
    filter(
      !is.na(.data[[column_name]])
    ) %>%
    mutate(
      classification = if_else(
        .data[[column_name]] >= threshold,
        "Above threshold",
        "Below threshold"
      )
    )
  
  return(cleaned_data)
}


```

# ------------------------------------------------------------
# 15. Create Mock Survey Data
# ------------------------------------------------------------

```{r}
survey_data <- tibble(
  site = c(
    "Site_A",
    "Site_B",
    "Site_C",
    "Site_D",
    "Site_E"
  ),
  coral_cover = c(
    42,
    NA,
    18,
    65,
    30
  )
)

survey_data

```
# ------------------------------------------------------------
# 16. Test the Data-Wrangling Function
# ------------------------------------------------------------

cleaned_survey <- clean_and_classify_survey(
  data = survey_data,
  column_name = "coral_cover",
  threshold = 30
)

cleaned_survey









